Top-down versus bottom-up cohesiveness in microbial community coalescence

### 2021-05-13 draft/draft_v2.tex
box 1 added

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{upgreek} % upright greek letters for math
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{makecell} % more table customization
\usepackage{mathtools} % nice matrices
\usepackage{units} % nice fractions
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\zxrsetup{toltxlabel}
\zexternaldocument*{supp}
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[table,xcdraw]{xcolor} % text and table background colors
\usepackage[margin=3cm]{geometry} % margins
\usepackage{lineno} % line numbers
\usepackage{textcomp} % special characters (some references break if this isn't loaded?)

\usepackage{pdflscape} %% Landscape
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{savesym}
\savesymbol{iint}
\restoresymbol{TXF}{iint}
%% \usepackage{pdfpages} %% to include a pdf file

@@

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness
in microbial community coalescence}
\author[1,2,*]{Juan Diaz-Colunga}
\author[1,2,*]{Nanxi Lu}
\author[1,2,*]{Alicia Sanchez-Gorostiaga}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1,2,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology,
Yale University, New Haven, CT, USA}
\affil[2]{Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[3]{Cell Metabolism, Agios Pharmaceuticals, Inc.,
Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

@@

Our results indicate that co-selection varies in direction and strength depending on the supplied limiting
resource.
This primary resource, in turn, has been shown to
shape the structure and interactions of the communities \cite{Estrela2020}.
We observe that, when top-down co-selection is weak, bottom-up co-selection can be very strong,
with positive co-selection being far more common than negative co-selection.
\iffalse
Our results confirm that positive top-down co-selection is common, but its effects are weak.
In contrast, bottom-up co-selection can be very strong, and positive co-selection is far more common
than negative co-selection.
\fi
We then turn to a microbial consumer-resource model (microCRM)
\cite{Goldford2018,Marsland2019,Marsland2020}
that is able to capture the dynamics of microbial communities dominated by metabolic interactions,
as is the case for the ones assembled in our experimental conditions \cite{Goldford2018,Estrela2020}.
We show that the empirically observed trends in ecological co-selection are reproduced
with minimal model assumptions, and that
tuning the complexity of the metabolic interactions in our \textit{in silico} communities can modulate
the recurrence of top-down or bottom-up co-selection.
Our findings indicate that collective interactions play an important role at dictating community structure
during coalescence.

@@

If communities being coalesced were highly cohesive from the top-down,
the dominant species would co-select the rarer members of its community during coalescence
(\figref[e]{fig1}, left panels).
In this scenario, we would expect the outcome of community coalescence to be predicted by
which of the two dominants is most competitive in pairwise competition.
Analogously, competition between dominants should be affected only weakly by the presence or
absence of the cohorts, that would play a passive role under these conditions.
To test this hypothesis, we performed all pairwise competitions between dominant species
in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
the cultures for seven serial transfers, roughly 42 generations
(\methodsref{competitions}).
We then performed all possible pairwise community coalescence experiments by mixing
equal volumes of the communities and propagating the resulting cultures for seven extra
transfers
(\figref[f]{fig1}),
and we quantified coalescence outcomes using the similarity between the coalesced and the
invasive communities
(\methodsref{metrics}).
The frequencies of all species in both community-community and dominant-dominant competitions
were determined by 16S Illumina sequencing
(\methodsref{sequencing}).

\mr Rearrange this section to follow the flow of the new figure 2.
Also introduce the model here and cook a nice experiment-theory-experiment sandwich later.

We found that, for communities assembled in the glutamine environment,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
but this correlation was stronger for the citrate communities
(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
Alternative quantifications of community distance yielded similar results,
with weaker effects when the metric used
accounted only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).
Furthermore, we observed that, in the glutamine communities,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. after coalescence
(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
On the other hand, pairwise competition between dominants of the citrate communities
is highly correlated with the relative abundance of said dominants in community coalescence
 (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).

Together, these results suggest that the strength of top-down co-selection depends on
the primary resource supplied to the coalesced communities.
Communities assembled with citrate as the primary supplied resource
display a strong degree of top-down co-selection,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to be passively responding to their dominants.
Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
This finding led us to investigate the potential role of bottom-up ecological co-selection
(\figref[e]{fig1}, right panels), i.e. whether in some cases the dominants may be co-selected
for or against by their cohorts.

@@

\subsubsection*{Simulations of community coalescence with a consumer-resource model}

To better understand the mechanisms of ecological co-selection in our experiments,
we ran a set of simulations of community coalescence. We used a microbial consumer-resource
model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be
sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020},
for which the microCRM provides a good description.
We implemented a new feature in the package that makes individual species
able to secrete different sets of byproducts when consuming a same resource
(\methodsref{sim}).
We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
types of byproducts when consuming a resource. These assumptions are supported by the experimental
observation that individual species can indeed secrete different sets of few metabolites
to the environment when consuming a primary resource
\cite{Harcombe2014,Pinu2018}. \mr cite Sylvie here?
We reason that these choices are necessary to potentially generate community cohesiveness:
if the secretions of all species were roughly similar, the effect of interchanging members of
a community would be minimal. But in practice, species with a history of coexistence make up
cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
In short,
we reproduced our experimental protocol \textit{in silico}
by generating a library of species that we divided into two pools. Each pool was used to seed a
collection of invasive and resident communities, that were allowed to stabilize through 20
growth-dilution cycles. We then mixed pairs of these stable communities to carry out our coalescence
and dominant-dominant competition experiments,
as well as the invasion experiments from the dominants alone.
A full description of our modeling framework can be found in the \methodsref{sim} section.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=15cm,keepaspectratio]{figs/fig2_v3.pdf}
\caption{\textbf{Top-down co-selection in microbial community coalescence.}
\textbf{a.} Pairwise competition of dominants with or without their cohorts.
In the horizontal axis, we plot the frequency of the invasive dominant species
in head-to-head pairwise competition with the resident dominant.
In the vertical axis, we plot the same relative frequency when the two species compete
in the presence of their cohorts, i.e. during community coalescence.
$R^2=0.04$, $p>0.05$ for glutamine and $R^2=0.83$, $p<10^{-8}$ for citrate.
\textbf{b.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity
($Q$) between the coalesced and invasive communities.
These outcomes are predicted by the pairwise competition between the invasive and
resident dominant species.
Left panel (red): glutamine communities, $R^2=0.15$, $p<0.05$.
Middle panel (blue): citrate communities, $R^2=0.57$, $p<10^{-4}$.
A high correlation is consistent with a scenario of
strong top-down positive co-selection where dominants recruit their cohorts for the final
coalesced community. Two biological replicates per experiment are plotted individually.
Right panel (black): simulations with a microbial consumer-resource model are able
to capture these trends ($R^2=0.22$, $p<10^{-5}$).}
\label{fig2}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=12cm,keepaspectratio]{figs/fig3.pdf}
\caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
\textbf{a.} Blablabla.
\textbf{b.} Blablabla.
\textbf{c.} Blablabla.
\textbf{d.} Blablabla.
\textbf{e.} Blablabla.
\textbf{f.} Blablabla.}
\label{fig3}
\end{figure}

>>>>

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{upgreek} % upright greek letters for math
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{makecell} % more table customization
\usepackage{mathtools} % nice matrices
\usepackage{units} % nice fractions
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\zxrsetup{toltxlabel}
\zexternaldocument*{supp}
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[table,xcdraw]{xcolor} % text and table background colors
\usepackage[margin=3cm]{geometry} % margins
\usepackage{lineno} % line numbers
\usepackage{textcomp} % special characters (some references break if this isn't loaded?)
\usepackage{mdframed} % to easily handle boxes

@@

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness
in microbial community coalescence}
\author[1,*]{Juan Diaz-Colunga}
\author[1,*]{Nanxi Lu}
\author[1,2,*]{Alicia Sanchez-Gorostiaga}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology
and
Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[2]{Department of Microbial Biotechnology,
Centro Nacional de Biotecnología (CNB-CSIC), Cantoblanco, Madrid, Spain}
\affil[3]{Physics of Living Systems, Department of Physics,
Massachusetts Institute of Technology, Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

@@

Our results indicate that co-selection varies in direction and strength depending on the supplied limiting
resource.
This primary resource, in turn, has been shown to
shape the structure and interactions of the communities \cite{Estrela2020}.
We observe that, when top-down co-selection is weak, bottom-up co-selection can be very strong,
with positive co-selection being far more common than negative co-selection.
\iffalse
Our results confirm that positive top-down co-selection is common, but its effects are weak.
In contrast, bottom-up co-selection can be very strong, and positive co-selection is far more common
than negative co-selection.
\fi
We then turn to a Microbial Consumer-Resource Model (MicroCRM)
\cite{Goldford2018,Marsland2019,Marsland2020}
that is able to capture the dynamics of microbial communities dominated by metabolic interactions,
as is the case for the ones assembled in our experimental conditions \cite{Goldford2018,Estrela2020}.
We show that the empirically observed trends in ecological co-selection are reproduced
with minimal model assumptions, and that
tuning the complexity of the metabolic interactions in our \textit{in silico} communities can modulate
the recurrence of top-down or bottom-up co-selection.
Our findings indicate that collective interactions play an important role at dictating community structure
during coalescence.

@@

If communities being coalesced were highly cohesive from the top-down,
the dominant species would co-select the rarer members of its community during coalescence
(\figref[e]{fig1}, left panels).
In this scenario, we would expect the outcome of community coalescence to be predicted by
which of the two dominants is most competitive in pairwise competition.
Analogously, competition between dominants should be affected only weakly by the presence or
absence of the cohorts, that would play a passive role under these conditions.
To test this hypothesis, we performed all pairwise competitions between dominant species
in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
the cultures for seven serial transfers, roughly 42 generations
(\methodsref{competitions}).
We then carried out all possible pairwise community coalescence experiments by mixing
equal volumes of the communities and propagating the resulting cultures for seven extra
transfers
(\figref[f]{fig1}).
\iffalse
and we quantified coalescence outcomes using the similarity between the coalesced and the
invasive communities
(\methodsref{metrics}).
\fi
The frequencies of all species in both community-community and dominant-dominant competitions
were determined by 16S Illumina sequencing
(\methodsref{sequencing}).

We found that, for communities assembled in the glutamine environment,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. during community coalescence
(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
This correlation is significantly higher for the citrate communities
 (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).
This suggests that, in the glutamine environments, head-to-head competition of dominants
is heavily influenced by higher order effects introduced by the rare taxa of the communities.
On the other hand, the cohorts seem to play a more passive role in the citrate environments.
To test the effects of top-down co-selection at the community level,
we quantified the distances between the invasive and coalesced communities
using the relative Bray-Curtis similarity (\methodsref{metrics})
and compared them to the outcomes of the pairwise competitions between dominants alone.
We again noticed differences between glutamine and citrate communities:
for the former,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(\figref[b]{fig2} left panel, $R^2=0.15$, $p<0.05$).
For the latter,
the structure of the coalesced communities tends to be more strongly dictated by the
result of the dominant-dominant competition
(\figref[b]{fig2} middle panel, $R^2=0.57$, $p<10^{-4}$).
Alternative quantifications of community distance yield similar results,
with weaker effects when the metric used
accounts only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).

Together, these results suggest that the strength of top-down co-selection depends on
the environment where communities are assembled and coalescence takes place.
Communities stabilized with citrate as the primary supplied resource
display a strong degree of top-down cohesiveness,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to just be passively responding to their dominants.
Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
To investigate the determinants of top-down co-selection and the factors modulating its strength,
we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
Model (MicroCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020},
for which the MicroCRM provides a good description.
We carried out our simulations enabling a large variation across species (consumers) in terms of
their metabolic characterization.
This choice is discussed in \hyperref[box1]{Box 1}.

\begin{mdframed}
\internallinenumbers
\subsubsection*{Box 1: Community cohesiveness in a Microbial Consumer-Resource Model}
\label{box1}

The Microbial Consumer-Resource Model (MicroCRM) \cite{Goldford2018,Marsland2019,Marsland2020}
is a modeling framework based on the classic MacArthur's consumer resource model
\cite{MacArthur1970}.
It encodes the dynamics of a system with $S$ species and $M$ resources
in terms of a consumer preference matrix $\mathbf{c}$ and a metabolic matrix $\mathbf{D}$,
with an additional set of parameters controlling
the species maintenance costs ($m_i$ for species $i$),
the resource energy densities ($w_\alpha$ for resource $\alpha$),
the energy to growth rate conversion factor ($g_i$ for species $i$)
or the leakage fraction, i.e. the amount of energy lost as byproducts when a resource is consumed
($l_\alpha$ for resource $\alpha$).
The element $c_{i\alpha}$ of the consumer preference matrix
represents the uptake rate of resource $\alpha$ by
species $i$ (although the relationship between $c_{i\alpha}$ and the uptake rate can be more
complex in modeling scenarios that are not considered here,
see \cite{Goldford2018,Marsland2019,Marsland2020}).
The element $D_{\beta\alpha}$ of the metabolic matrix
represents the amount of energy secreted in the form of resource $\beta$
as a result of the metabolism of resource $\alpha$.

This formulation assumes that all species metabolize resources through similar pathways,
with any two species secreting the same byproducts when consuming a same resource.
However, experimental evidence suggests that individual species can secrete different sets of metabolites
to the environment when growing on a same primary resource
\cite{Harcombe2014,Pinu2018,Estrela2020}. This observation motivated us to introduce a new
feature in the model:
in short, we now define $\mathbf{D}$ as a three-dimensional matrix
where the element $D_{i\beta\alpha}$ represents the energy flux in the form of resource $\beta$
that is secreted \textit{by species $i$} when it metabolizes resource $\alpha$.
Note that now $D_{i\beta\alpha}$ need not be equal to $D_{j\beta\alpha}$ if $i \neq j$
(see illustration below).

\bigskip
\begin{center}
\includegraphics[width=8cm,keepaspectratio]{figs/figBox.png}
\end{center}
\bigskip

We reason that the choice of species-specific metabolic fluxes is necessary to potentially generate
cohesiveness at the community level during coalescence.
If the secretions of all species were identical (or only slightly different),
higher order cross-feeding effects would be very unspecific:
the establishment of new invasive species
--given that they could outcompete resident taxa within their metabolic niches,
i.e. more effectively feed off the same resource or set of resources as them--
would not alter (or only do so moderately)
the metabolic flows through the rest of the cross-feeding network.
On the other hand, said network could undergo a profound and further-reaching restructuring
if the invasive species secreted very different sets of metabolites with respect to the resident ones,
potentially disabling existing niches and/or enabling new ones
where more invaders could be co-selected.
For a similar reason, we argue that the sparsity of the metabolic matrix could also modulate
the emergence of cohesiveness in the face of coalescence.
A dense metabolic matrix corresponds to a situation where all species secrete a wide
variety of byproducts.
New-coming invasive species that secrete similar byproducts as resident ones
--even if they do so in different amounts--
might only induce moderate quantitative changes in the metabolic fluxes.
But if the sets of secretions are qualitatively different, co-selection of species adapted to
each of those sets becomes possible.
These ideas are supported by experimental observations
suggesting that species with a history of coexistence make up
cohesive communities with highly specific cross-feeding configurations
\cite{Rosenzweig1994,Goldford2018,Estrela2020}.

\bigskip
\end{mdframed}

\iffalse
We found that, for communities assembled in the glutamine environment,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
but this correlation was stronger for the citrate communities
(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
Alternative quantifications of community distance yielded similar results,
with weaker effects when the metric used
accounted only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).
Furthermore, we observed that, in the glutamine communities,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. after coalescence
(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
On the other hand, pairwise competition between dominants of the citrate communities
is highly correlated with the relative abundance of said dominants in community coalescence
 (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).

Together, these results suggest that the strength of top-down co-selection depends on
the primary resource supplied to the coalesced communities.
Communities assembled with citrate as the primary supplied resource
display a strong degree of top-down co-selection,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to be passively responding to their dominants.
Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.

This finding led us to investigate the potential role of bottom-up ecological co-selection
(\figref[e]{fig1}, right panels), i.e. whether in some cases the dominants may be co-selected
for or against by their cohorts.
\fi

@@

\subsubsection*{Simulations of community coalescence with a Consumer-Resource Model}

To better understand the mechanisms of ecological co-selection in our experiments,
we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
Model (MicroCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be
sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020},
for which the MicroCRM provides a good description.
We implemented a new feature in the package that makes individual species
able to secrete different sets of byproducts when consuming a same resource
(\methodsref{sim}).
We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
types of byproducts when consuming a resource. These assumptions are supported by the experimental
observation that individual species can indeed secrete different sets of few metabolites
to the environment when consuming a primary resource
\cite{Harcombe2014,Pinu2018}. \mr cite Sylvie here?
We reason that these choices are necessary to potentially generate community cohesiveness:
if the secretions of all species were roughly similar, the effect of interchanging members of
a community would be minimal. But in practice, species with a history of coexistence make up
cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
In short,
we reproduced our experimental protocol \textit{in silico}
by generating a library of species that we divided into two pools. Each pool was used to seed a
collection of invasive and resident communities, that were allowed to stabilize through 20
growth-dilution cycles. We then mixed pairs of these stable communities to carry out our coalescence
and dominant-dominant competition experiments,
as well as the invasion experiments from the dominants alone.
A full description of our modeling framework can be found in the \methodsref{sim} section.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=15cm,keepaspectratio]{figs/fig2_v3.pdf}
\caption{\textbf{Top-down co-selection in microbial community coalescence.}
\textbf{a.} Pairwise competition of dominants with or without their cohorts.
In the horizontal axis, we plot the frequency of the invasive dominant species
in head-to-head pairwise competition with the resident dominant.
In the vertical axis, we plot the same relative frequency when the two species compete
in the presence of their cohorts, i.e. during community coalescence.
$R^2=0.04$, $p>0.05$ for glutamine and $R^2=0.83$, $p<10^{-8}$ for citrate.
\textbf{b.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity
($Q$) between the coalesced and invasive communities.
These outcomes are predicted by the pairwise competition between the invasive and
resident dominant species.
Left panel (red): glutamine communities, $R^2=0.15$, $p<0.05$.
Middle panel (blue): citrate communities, $R^2=0.57$, $p<10^{-4}$.
A high correlation is consistent with a scenario of
strong top-down positive co-selection where dominants recruit their cohorts for the final
coalesced community. Two biological replicates per experiment are plotted individually.
Right panel (black): simulations with a Microbial Consumer-Resource Model are able
to capture these trends ($R^2=0.22$, $p<10^{-5}$).}
\label{fig2}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=12cm,keepaspectratio]{figs/fig3.pdf}
\caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
\textbf{a.} We hypothesize that three scenarios are possible regarding bottom-up co-selection:
sub-dominant species could co-select for (green) or against (red) their dominant in coalescence,
or they could have no effect in the invasion success of the dominant taxa (gray).
\textbf{b.} Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
co-selection is common and can be very strong, whereas negative bottom-up co-selection is rare.
We plot the frequency reached by the invasive dominants when invading the resident communities
in isolation versus the same frequency when invading together with their cohorts,
i.e. in community coalescence.
Points in the green/red area represent instances where the invasive dominant is able to invade
with higher/lower success when accompanied by its cohort, evidencing positive/negative
bottom-up co-selection.
Points around the diagonal (gray area) correspond to cases where the success of the invasive dominant
is only weakly affected by the presence or absence of its cohort.
\textbf{c.} We divided the data from our simulations into two sets according to whether
positive or no bottom-up co-selection was observed (that is, whether points fell into the green or gray
areas of panel b). Here we reproduce the plots in \figref[b]{fig2} for each set, representing the
result of the dominant head-to-head pairwise competition versus the outcome of community coalescence.
Left panel: when positive bottom-up co-selection is strong, head-to-head pairwise competition of dominants
is poorly predictive of coalescence outcomes
($R^2=0.00$, $p>0.05$).
Right panel: on the other hand, when bottom-up co-selection is weak coalescence outcomes are more
strongly dictated by the result of the dominant-dominant competition
($R^2=0.34$, $p<10^{-6}$).
\textbf{d.} Experiments show that in our conditions, positive bottom-up co-selection is indeed more
frequent and strong than negative bottom-up co-selection.
\textbf{e.} We reproduce the plots in panel c for our experimental data. This is equivalent to recreating
\figref[b]{fig2}, this time splitting our data by the strength of bottom-up co-selection
instead of by the carbon source provided to the communities.
Like simulations predicted, instances of community coalescence displaying strong positive bottom-up co-selection
yield a poor correspondence between dominant pairwise competition and coalescence outcomes
(left panel, $R^2=0.07$, $p>0.05$)
and vice-versa
(right panel, $R^2=0.37$, $p<10^{-4}$).}
\label{fig3}
\end{figure}
